The E-C Algorithm: conceptual bootstrapping via grammar-based program compression.

### 2013-01-18 drafts/ijcai2013/ijcai13.tex
Changes.

\usepackage{ijcai13}
\usepackage{times}
\usepackage{amsmath} 
\usepackage{latexsym} 
\usepackage{ dsfont }
\usepackage{tikz}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usetikzlibrary{arrows,automata}

\DeclareMathOperator*{\argmax}{arg\,max}

\title{The Exploration-Compression Algorithm: how to learn programs by discovering useful subroutines.}
\author{Eyal Dechter \\
MIT\\
USA \\
edechter@mit.edu
\And
Josh Tenenbaum \\
MIT\\
USA \\
jbt@mit.edu
\And 
Ryan Adams \\
MIT\\
USA \\
rpa@seas.harvard.edu}

@@

\begin{abstract}
  Every programmer knows that the abstractions provided by a good
  function library are not merely a convenience: they usefully
  constrain the space of implementations and guide the choice of
  programming patterns. Without these abstractions, whether explicit
  or implicit, the designer would wander in a sea of
  detail. Similarly, the learner who wishes to discover the theories
  and concepts that explain the world must use a set of abstractions
  to guide their development. But how do those abstractions arise?
  Here, we show how alternating between a search over program space
  and a compression of discovered programs allows a learner to build
  up abstractions that make the learner's search progressively more
  effective.
\end{abstract}

@@

@@

Following~\cite{liang10programs} and~\cite{Briggs:2008}, we use a
combinatory logic -- a variable-free subset of the polymorphic
simply-typed lambda calculus lambda calculus~\cite{Pierce_2002} -- as
our program representation language. In short, the combinatory logic
introduces a \emph{basis} of several primitive \emph{combinators} such
that any function in the lambda calculus can be alternatively written
as applications of the basis combinators. The basis combinators can
themselves been expressed in the lambda calculus. The lambda calculus
has two basic operation -- application and abstraction -- but in using
the combinatory logic we sequester uses of the abstraction operation
inside the combinators, and thus have a representation language that
is effectively abstraction- and, thus, variable-free.

This is very convenient for program synthesis: since every expression
is the application of one expression to another -- with this recursion
bottoming out at the primitive combinators -- each program is a binary
tree. Most importantly, any subtree is itself a well-formed
expression; this is not the case in the lambda calculus, since
abstraction introduces long range dependencies between the $lamdba$-
operator and the variables to which that operator refers. In the
lambda calculus, then, a subtree might have free variables, not bound
to any enclosing $lambda$.

\subsection{Enumerating the most likely programs.}
The simplicity of our program representation suggests a simple
recursive recipe for generating all programs of a given type $\tau$ up
to some maximal depth $d$: such a program is either one of the atomic
programs with type in $\tau$, or it is pairing of a left child program
and a right child program with appropriate types each of which comes
from corresponding sets of programs with depth up to $d-1$. The left
child is any program with type $\sigma' \leftarrow \tau'$ which is a
subset of $\sigma \leftarrow \tau$, where $\sigma$ is a free type
variable. To match, the right child must have type $\sigma'$. The
resulting program will have type $\tau'$ which is a subset of $\tau$.

  \begin{algorithm}
    \SetAlgoLined

    \KwData{Library of primitive/type pairs: $\mathcal{L} = \{(f_1, \tau_1), \dots,
        (f_m, \tau_m)\}$. Target type $\tau^*$. Max depth $d$.}

    Choose $F$ from $\text{ ENUM}(\mathcal{L}, \tau^*, 0)$, or:

    Choose $L$ from $\text{ ENUM}( \mathcal{L}, \sigma \rightarrow
    \tau^*, d-1)$.

    Choose $R$ from $\text{ ENUM}( \mathcal{L}, sourceType(L), d-1)$.

    Return $F = (L\,R)$ with type $targetType(L)$.

    \caption{ENUM}
  \end{algorithm}

\subsection{Finding the most compressive solutions.}

\subsection{Re-estimating the probabilistic grammar over programs.}

\section{Experiments.}
\subsection{Integers: a toy case study.}
\subsection{Symbolic regression.}
\subsection{Learning electric circuits.}
\begin{figure}[h]
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2cm,
                    , minimum size=.5cm, thick,main node/.style={circle,draw}]

  \node[main node] (S) {$\{\}$};
  \node[main node] (1) [below left of=S] {$\{a\}$};
  \node[main node] (2) [below right of=S] {$\{b\}$} ;
  \node[main node] (3) [below left of=1] {$\{a, a\}$};
  \node[main node] (4) [below right of=1] {$\{a, b\}$};
  \node[main node] (5) [below right of=2] {$\{b, b\}$};;

  \path
    (S) edge node [left] {a} (1)
        edge node [right] {b} (2)
    (1) edge node [left] {a} (3)
        edge node [right] {b} (4)
    (2) edge node [left] {a} (4)
        edge node [right] {b} (5)

\end{tikzpicture}
\end{figure}

>>>>

\usepackage{ijcai13}
\usepackage{times}
\usepackage{amsmath} 
\usepackage{latexsym} 
\usepackage{ dsfont }
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usetikzlibrary{arrows,automata, positioning}

\newtheorem{definition}{Definition}
\DeclareMathOperator*{\argmax}{arg\,max}

\title{The E-C Algorithm: conceptual bootstrapping via grammar-based
  program compression.}

\author{Eyal Dechter \\
MIT\\
USA \\
edechter@mit.edu
\And
Josh Tenenbaum \\
MIT\\
USA \\
jbt@mit.edu
\And 
Ryan Adams \\
Harvard University\\
USA \\
rpa@seas.harvard.edu}

@@

\begin{abstract}
Suppose a learner is faced with a set of problems about which it knows
nearly nothing: it doesn't know the distribution of problems, or what
an appropriate loss function might be. For each hypothesized solution
to a problem it gets a few bits of positive or negative reinforcement,
but not enough to steer it effectively in the huge space of possible
solutions. How can such a learner every get off the ground? The key is
to find, by brute force, a solution to one or more simple problems in
that domain and use the information thus gained to pursue subsequent
searches in an informed manner. Here, we formalize this intuition,
where the solution space is the space of typed functional programs and
the gained information is stored as a stochastic grammer over
programs. Specifically, we propose the E-C algorithm, a two step
iterative procedure for exploring such spaces: in the E step, the
learner \emph{explores} a finite subset of the domain for solutions,
guided by a stochastic grammar; in the C step, the learner
\emph{compresses} the successful solutions from the E step to estimate
a new stochastic grammar. As we show in several toy domains, iterating
these two steps quickly propels the learner from a state of solving
almost no problems to solving a large number of problems by gaining
abstract knowledge of the structure of the solution space.
\end{abstract}

@@

Suppose that, having been shipwrecked and saved by a people with no
knowledge of your language or culture, you find yourself attempting to
communicate in a language about which you know nothing. How would you
go about discovering even the most basic mappings of sounds to
meanings? Overhearing the native speech, you have a sense of what the
phonetics might be like, so you might try various common combinations
of sounds while pointing to things you want. Mostly, you receive
puzzled looks, but if you are lucky, you might pick up on a few key
words or phrases, like ``that'' or ``I want.'' Naturally, you will
proceed to use your discoveries to accelerate your learning, mixing
and matching known phrases to generate new meanings.

@@

Following~\cite{liang10programs} and~\cite{Briggs:2008}, we use
combinatory logic -- a variable-free subset of the polymorphic
simply-typed lambda calculus lambda calculus~\cite{Pierce_2002} -- as
our program representation language. In short, the combinatory logic
introduces a \emph{basis} of several primitive \emph{combinators} such
that any function in the lambda calculus can be alternatively written
as applications of the basis combinators. Some common basis
combinators are defined as follows:
\begin{align}
I\; x &\rightarrow  x \\
S\; f\; g\; x\; &\rightarrow (f\; x)\; (g\; x) \\ 
B\; f\; g\; x\; &\rightarrow (f\; x)\; g \\ 
C\; f\; g\; x\; &\rightarrow f\; (g\; x) \\ 
\end{align}

The basis combinators can
themselves been expressed in the lambda calculus. The lambda calculus
has two basic operation -- application and abstraction -- but in using
the combinatory logic we sequester uses of the abstraction operation
inside the combinators, and thus have a representation language that
is effectively abstraction- and, thus, variable-free.

This is very convenient for program synthesis: since every expression
is the application of one expression to another -- with this recursion
bottoming out at the primitive combinators -- each program is a binary
tree. Most importantly, any subtree is itself a well-formed
expression; this is not the case in the lambda calculus, since
abstraction introduces long range dependencies between the $\lambda$
operator and the variables to which that operator refers. In the
lambda calculus, then, a subtree might have free variables, not bound
to any enclosing $\lambda$.

As a simple example of our representation, consider the squaring
function $\lamba x. * x x $. Using two of the basis combinators above,
we can write the squaring function in combinatory logic as $S * I$,
where application associates to the left, so we drop the
parentheses. When we apply this combinator to a value $x$, the action
of the combinator is defined as $S \;*\; I\; x~\rightarrow~(*\; x)\;
(I\; x)~\rightarrow(*\; x)\; x$. Now, we can extend this
representation with a simple polymorphic type
system~\cite{Pierce_2002}. Thus, this combinator can be represented as
a binary tree~(Figure~\ref{fig:clbintree}) whose leaves are typed
primitive combinators and whose interior nodes are typed applications of combinators.

\begin{figure}
  \centering
  \begin{subfigure}{0.2\linewidth}
    \label{fig:clbintree}
    \begin{tikzpicture}[scale=.5, transform shape, level/.style={sibling distance=10mm/#1, font=\scriptsize}]
      \node [font=\scriptsize] (top) {$I$}
      child {node [above=.3cm] (a) {$(I \rightarrow I) \rightarrow I \rightarrow I$}
        child {node [left=.5cm, below=1cm, text width=3cm] 
          {$S : \\(t_2 \rightarrow t_1 \rightarrow t_0)  \rightarrow (t_2 \rightarrow t_1) 
            \rightarrow \\ t_2 \rightarrow t_0$}}
        child {node [right=0cm, text width=2cm] {$*: Int \rightarrow Int \rightarrow Int$}}
      }
      child {node [below right=.3cm and 1cm] (b) {$I \rightarrow I$}
        child {node {$I: t_0 \rightarrow t_0 \rightarrow t_0$}}
      };

    \end{tikzpicture}
  \end{subfigure}

  \begin{subfigure}[b]{0.2\linewidth}
    \label{fig:andor}
    \begin{tikzpicture}[scale=.5, transform shape, level/.style={sibling distance=10mm/#1}]
      \node [circle,draw] (a) {}
      child {node [left=0cm] (b) {$f_1$}}
      child {node [left=.2cm] (c) {\dots} edge from parent[draw=none] }
      child {node [left=.4cm] (e) {$f_n$}}
      child {node [circle,draw, left=1cm] (d) {}
        child {node [circle, draw, left=.25cm] (l) {} edge from parent[draw=none]
          child {node [left=.5cm] (b) {$f_1$}} 
          child {node [left=.3cm] (c) {\dots} edge from parent[draw=none] }
          child {node [left=.3cm] (e) {$f_n$}}
          child {node [circle, draw, left=.25cm] (and2) {}
            child {node [circle, draw, left=.25cm](l2) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw] (r2)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
        }
        child {node [circle, draw, right=.25cm] (r) {} edge from parent[draw=none]
          child {node [circle, draw, right=.25cm] (and3) {}
            child {node [circle, draw ](l3) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw, right=.2cm ] (r3)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
          child {node [right=.3cm] (b) {$f_1$}} 
          child {node [right=.4cm] (c) {\dots} edge from parent[draw=none] }
          child {node [right=.6cm] (e) {$f_n$}}
        }
      };

      \path (d)  edge node[inner sep=0mm,pos=.4] (a1) {} node {} (l);
      \path (d)  edge node[inner sep=0mm,pos=.4] (b1) {} node {} (r);
      \path (a1) edge[bend right] (b1) ;

      \path (and2)  edge node[inner sep=0mm,pos=.4] (t0) {} node {} (l2);
      \path (and2)  edge node[inner sep=0mm,pos=.4] (k0) {} node {} (r2);
      \path (t0) edge[bend right] (k0) ;

      \path (and3)  edge node[inner sep=0mm,pos=.4] (t1) {} node {} (l3);
      \path (and3)  edge node[inner sep=0mm,pos=.4] (k1) {} node {} (r3);
      \path (t1) edge[bend right] (k1) ;
    \end{tikzpicture}

    \end{subfigure}
  \caption{A typed binary tree representation of the combinator $S\;*\;I\;$.}
\end{figure}

\subsection{Best-first enumeration of most probably programs.}
In order to explore the frontier of most probable programs, we need a
procedure that enumerates the $N$ most probable programs. We can
formulate this naturally as a best-first exploration of an $AND-OR$
tree (NEED FIGURE). Complete programs are represented by binary trees
that are valid subtrees of the $AND-OR$ graph. The tree structure is
shown in Figure~\ref{fig:andor}. The root is an $OR$ node of the
requisite program type. It's children are the primitives of that type
and one $AND$ node with two children. Each child has the same
structure as the root node, except that the types are different. If
the type of an $OR$ node is $t$, then the type of the left child of
its $AND$ node child is $\tau \rightarrow t$ (where $\tau$ is a fresh
type variable). Since the type of the right child is constrained by
the subtree rooted at the left child, we always expand the left child
first, and we use the target type of this left subtree to mark the
type of the right child.

\begin{figure}
\label{fig:andor}
\begin{tikzpicture}[scale=.5, transform shape, level/.style={sibling distance=10mm/#1}]
\node [circle,draw] (a) {}
child {node [left=0cm] (b) {$f_1$}}
child {node [left=.2cm] (c) {\dots} edge from parent[draw=none] }
child {node [left=.4cm] (e) {$f_n$}}
child {node [circle,draw, left=1cm] (d) {}
  child {node [circle, draw, left=.25cm] (l) {} edge from parent[draw=none]
    child {node [left=.5cm] (b) {$f_1$}} 
    child {node [left=.3cm] (c) {\dots} edge from parent[draw=none] }
    child {node [left=.3cm] (e) {$f_n$}}
    child {node [circle, draw, left=.25cm] (and2) {}
          child {node [circle, draw, left=.25cm](l2) {} edge from parent[draw=none] 
            child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
          }
          child {node [circle, draw] (r2)  {} edge from parent[draw=none] 
            child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
          }
    }
  }
  child {node [circle, draw, right=.25cm] (r) {} edge from parent[draw=none]
    child {node [circle, draw, right=.25cm] (and3) {}
          child {node [circle, draw ](l3) {} edge from parent[draw=none] 
            child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
          }
          child {node [circle, draw, right=.2cm ] (r3)  {} edge from parent[draw=none] 
            child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
          }
    }
    child {node [right=.3cm] (b) {$f_1$}} 
    child {node [right=.4cm] (c) {\dots} edge from parent[draw=none] }
    child {node [right=.6cm] (e) {$f_n$}}
 }
};

\path (d)  edge node[inner sep=0mm,pos=.4] (a1) {} node {} (l);
\path (d)  edge node[inner sep=0mm,pos=.4] (b1) {} node {} (r);
\path (a1) edge[bend right] (b1) ;

\path (and2)  edge node[inner sep=0mm,pos=.4] (t0) {} node {} (l2);
\path (and2)  edge node[inner sep=0mm,pos=.4] (k0) {} node {} (r2);
\path (t0) edge[bend right] (k0) ;

\path (and3)  edge node[inner sep=0mm,pos=.4] (t1) {} node {} (l3);
\path (and3)  edge node[inner sep=0mm,pos=.4] (k1) {} node {} (r3);
\path (t1) edge[bend right] (k1) ;
\end{tikzpicture}
\end{figure}

In accordance with the stochastic grammar over programs, we weight all
edges from $OR$ nodes according to the negative log likelihood of
choosing the associated rule from the stochastic grammar, conditioned
on the requesting type. Expanding an $AND$ node has zero cost. In the
case that the stochastic grammar is ill condition, we might have an
infinitely deep solution tree whose cost does not increase as we
unfold it; this can occur, for example, if there are no primitives of
the required type since the log probability of expansion to the next
$AND$ node is 0. . To avoid this, we simply declare a cost for
expanding the next $AND$ node when it is the only node available.

\subsection{Finding the most compressive set of solution.}

Once we have enumerated the best $N$ programs from our distribution,
we will find that some tasks have no solutions and some have one or
more solutions; for those tasks that have solutions, we want to choose
a solution for each task such that the complete set of such solutions
for all the tasks is as compressible as possible.

A natural metric for the compressibility of a set of binary trees is
the number of unique subtrees that this set contains. Thus, we can
attempt to find the assignment of solutions to tasks that minimizes
this value. However, an exact optimization for this problem requires
examining $O(m^k)$ assignments, where $m$ is the maximum number of
solutions for any one task and $k$ is the number of tasks. Since this
is prohibitive, we relax the compressibility metric as follows: we
define the cost of adding each new solution $s_n$ to a partial set of
solutions $s_1, \dots, s_{n-1}$ to be the number of additional unique
subtrees in $s_n$ that are not present in $s_{n-1}$. Our solution thus
becomes approximate and that approximation is order-dependent, but a
depth-first search on the defined search space goes from exponential
in the solution degree to quadratic.

\subsection{Re-estimating the stochastic grammar over programs.}

Given a set of solution programs, we want to re-estimate our
stochastic grammar. Our algorithm for doing this is based broadly on
the Neville-Manning algorithm for grammar-based
compression~\cite{nevill1997identifying}. In our algorithm, we simply
count the numer of times each unique subtree appears as the child of a
subtree \emph{not yet encountered} in the solution set. That is, we
traverse the solution set in depth first order, comparing each tree we
encounter to all the trees we previously encountered; but if we
encounter a tree we have already seen, we do not descend into it. In
this traversal, the set of trees we encounter more than once is the
set of terminal nodes in our new grammar. To estimate the
probabilities associated with each node production, we again traverse
the solution set, but these time for each node where a terminal
production takes place we count which other terminal productions could
have taken place. We estimate the terminal production weight to be the
ratio of the number of times that node was used in the first traversal
divided by the number of times it was used in the second traversal
(plus 1, to account for the event of an $AND$ node
expansion). Likewise, we re-estimate the probability of an $AND$ node
expansion by diving the number of such expansions in the solution set
by the number of possible expansions in the solution set.

\section{Experiments.}
\subsection{Symbolic regression.}

In this section, we walk through the performance of our learning
algorithm on a symbolic regression problem. In this problem,
each task is a set of pairs $ (x, f(x))$ where $x \in (0, \dots, 10)$
and $f$ is a polynomial. The goal is to discover a program for each
task that outputs the appropriate value $f(x)$ for each input x. We
assume that the data is noise-free.

The choice of an initial grammar for the algorithm will, of course,
have a big effect on its performance, and so we would like to
characterise this effect. We can divide primitives into roughly two
sets; there are low-level primitives, those that operate on concrete
types like Bool or Double, and these tend to be domain specific. For
example, conrete primitives include $1, 2$, etc., and $f(x) = 3x$. The
other class of primitives are abstract primitives, which can be
characterized by having arguments whose types are primarily function
types, like $(t \rightarrow s), (t \rightarrow s) \rightarrow \text{
  Double },$ etc. Of course, we can construct operators that fall in
between.

The most abstract operators are those that Liang et. al refer to as
``routers''~\cite{liang10programs}. These operators take the place of
$\lambda$ abstraction in the general lambda calculus: whereas
$\lambda$ abstraction allows us to bind arguments to variables nested
deep inside an expression, without them we need to have operators that
``route'' arguments to the appropriate functions in a nested
expression.

\begin{definition}[Router~\cite{liang10programs}] A \emph{router} is a combinator
represented by a finite sequence of elements from $\{B, C, S\}$. The
behavior of router $r$ is given by
\[
(r \, f \, g \, x_1 \dots x_{|r|}) = (f \, x_{i_1} \dots x_{i_n}) (g
\, x_{j_1} \dots x_{j_n}),
\]
where $i_1 < \dots < i_n$ are indices $i$ such that $r_i \in \{C, S\}$
and where $j_1 < \dots < j_n$ are indices $j$ such that $r_j \in \{B,
S\}$. Let $\mathcal{R}_k$ be the set of all routers of length equal to
$k$. Let $\mathcal{R}_{\leq k}$ be the set of all routers of length
less than or equal to $k$.
\end{definition}

To analyze the behavior of our algorithm, we will vary the parameters
of the problem along two axes while keeping the learning tasks the
same. Along the first axis, we vary the initial library to which our
learner has access with respect to the its abstractness (i.e., number
of routers) and its concreteness (i.e. number of concrete
primitives). Along the second axis we vary the size of the search
frontier.

Each task is to predict the value of an unknown polynomial on the
inputs $0, \dots, 10$. The set of tasks corresponds to the set of
polynomials $\{ax^2 + bx + c | a, b, c \in \{0, \dots, 10 \}\}$.

In Figure~\ref{fig:symreg}, we show performance results as we vary
both the routers included and the frontier size. Each individual plot
shows how changing the frontier size changes the number of tasks the
algorithm identifies correctly over the course of ten algorithm
iterations. In addition, we have a baseline for each curve, visits the
same number of expressions as the corresponding algorithm run
(i.e. frontier size $\times$ $\#$ of iterations) but all at once; it
does not do any learning. Thus, this baseline serves as a brute-force
control for the learning aspect of the algorithm.

\begin{figure}
\label{fig:symreg}
\includegraphics[width=\linewidth]{figures/learningCurves.pdf}
\includegraphics[width=\linewidth]{figures/performanceFrontierSize.pdf}
\caption{E-C learning curves for the symbolic regression task set as
  we vary the frontier size.}
\end{figure}

Why do the learning curves asymptote early? One might object that a
learning algorithm should continuously improve, if only slowly, as it
is given more time. But in these runs we observe a finite and
permanent performance cap. This is expected: once the learner find a
set of concepts it can explain, if those concepts fill up the
frontier, then there is nothing to spur the learner to incorporate as
yet unsolved tasks.

To investigate whether this is in fact the case, we can examine the
proportion of expressions in the frontier that match tasks in our task
set; that is, is the learning curve plateauing because the frontier is
heavily redundant, or is it plateuing because the frontier contains
many expressions that do not hit the task set at
all. Figure~\ref{fig:plateau_investigation} shows the proportion of
the expressions in the frontier that hit a task as the frontier sizeincreases.

\subsection{Digital arithmetic}
Having demonstrated the behavior of the E-C algorithm on symbolic
regression, we analyze its performance on tasks involving digital
arithmetic, a class of problems with known modular structure. The
general goal is to discover boolean circuits consisting of a limited
set of gates that perform arithmetic operations on binary number
representations.
